An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2011-03-17 paper.tex
Finished editing III.

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling \cite{Skilling:2006} and MultiNest \cite{Feroz:2009}).  However, we
frequently want to compute the posterior PDFs within each model along
with the evidences for the various models.  One of the most common
techniques for computing posterior PDFs in the context of a model is
the Markov chain Monte Carlo, which we now describe.

@@

Finally, we note that, in practice, the randomly chosen initial
starting point of the MCMC may be in a particularly unlikely location
in the parameter space.  Because jumps are frequently local, we will
generally want to ignore the early points in a finite-size chain to
avoid biases in the recovered posterior PDF due to the choice of the
initial location.  The points thus discarded are referred to as
"burn-in" points.

@@

The samples produced by an MCMC algorithm can be used to directly
perform a Monte Carlo evidence integral.  This results in a harmonic
mean estimator for the evidence, which suffers from a large variance
and bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.
Additional techniques for the direct integration of evidence, also
based on a kD tree decomposition of the parameter space (see
Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
techniques are promising, but in some cases suffer \cite{Farr2010}
from large variance and bias as well.  An alternative approach to
model selection among a set of models is based on performing an MCMC
in a ``super-model'' that encompasses all of the models under
consideration; this is known the the Reversible Jump Markov chain
Monte Carlo (RJMCMC).

@@

Ref.~\cite{Littenberg2009} decomposed parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each point was associated with the brick that
contained it.  Unfortunately, this process does not guarantee that
every brick contains at least one point, which is required for the
jump proposal to cover the entire model parameter space.  The solution
to this problem was to add a uniform distribution to the proposal PDF,
in the form of one ``fictitious'' point per brick.  The extra uniform
distribution from the fictituous points can represent a significant
cost, since inter-model jumps proposed from the uniform distribution
are unlikely to be accepted.  Additionally, the bricks in this
algorithm do not adapt to the local structure of the PDF.  One must
either use small bricks to capture the local structure of the PDF,
placing many bricks in regions without MCMC samples, or use large
bricks, missing the local strucutre of the PDF in exchange for fewer
empty bricks.

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions associated with each MCMC sample.  The
Voronoi region associated with a sample contains all the parameter
space points that are closer to that sample than any other sample.
The Voronoi region decomposition into neighborhoods is, in a sense,
maximally adaptive, in contrast to the approach of
Ref.~\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi regions requires a metric on
parameter space, which may be difficult or impossible to define.
Also, the computational cost for computing the Voronoi regions
increases rapidly with
dimension.  % Apparently the exact computational cost is not known in
            % arbitrary dimension.

Here we propose to use a decomposition of parameter space into
neighborhoods based on a datastructure called a kD-tree (see, e.g.\
\cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
efficient in high dimension than the Voronoi decomposition.

>>>>

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling \cite{Skilling:2004,Skilling:2006} and its variant MultiNest \cite{Feroz:2009}).  Although nested sampling can be used to compute the posterior PDFs within each model along
with the evidences for the various models, perhaps the most common
technique for computing posterior PDFs in the context of a model is
the Markov chain Monte Carlo, which we now describe.

@@

Finally, we note that, in practice, the randomly chosen initial
starting point of the MCMC may be in a particularly unlikely location
in the parameter space.  Because jumps are frequently local, we will
generally want to ignore the early points in a finite-size chain to
avoid biases in the recovered posterior PDF due to the choice of the
initial location.  The points thus discarded are referred to as
``burn-in'' points.

@@

The samples produced by an MCMC algorithm can be used to directly
perform a Monte Carlo evidence integral.  This results in a harmonic
mean estimator for the evidence, which suffers from a large variance
and bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.
Additional techniques for the direct integration of evidence, also
based on a kD tree decomposition of the parameter space (see
Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
techniques are promising, but in some cases suffer \cite{Farr2010} \will{Is this really the reference you meant?}
from large variance and bias as well.  An alternative approach to
model selection among a set of models is based on performing an MCMC
in a ``super-model'' that encompasses all of the models under
consideration; this is known the the Reversible Jump Markov chain
Monte Carlo (RJMCMC).

@@

Ref.~\cite{Littenberg2009} decomposed the parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each sample was associated with the brick that
contained it, and the probability of proposing a jump into a particular brick is thus proportional to the
number of samples within that brick.  Additionally, an extra uniform jump proposal was added to allow for jumps into bricks that did not contain any points, so that the jump 
proposal covers the entire model parameter space.  
%The solution to this problem was to add a uniform distribution to the proposal PDF, in the form of one ``fictitious'' point per brick.  The extra uniform distribution from the fictitious points can represent a significant cost, since inter-model jumps proposed from the uniform distribution are unlikely to be accepted.  
However, the bricks in this
algorithm do not adapt to the local structure of the PDF.  One must
either use small bricks to capture the local structure of the PDF,
placing many bricks in regions without MCMC samples (which can increase memory management and access costs, even for sparse matrices), or use large
bricks, missing the local structure of the PDF in exchange for fewer
empty bricks. \will{Are you happy with this bit of editing?}

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions associated with each MCMC sample.  The
Voronoi region associated with a sample contains all the parameter
space points that are closer to that sample than any other sample.
The Voronoi region decomposition into neighborhoods is, in a sense,
maximally adaptive, in contrast to the approach of
Ref.~\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi regions requires a metric on
parameter space, which may be difficult or impossible to define.
Also, the computational cost for computing the Voronoi regions
increases rapidly with dimensionality.  
% Apparently the exact computational cost is not known in arbitrary dimension.

Here we propose to use a decomposition of the parameter space into
neighborhoods based on a data structure called a kD-tree (see, e.g.\
\cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
efficient in high-dimensional spaces than the Voronoi decomposition.
